% !TeX root = ../main.tex
% Add the above to each chapter to make compiling the PDF easier in some editors.

\chapter{Introduction}\label{chapter:introduction}

Martingales hold a central position in the theory of stochastic processes, making them a fundamental concept for the working mathematician. They provide a powerful way to study and analyze random phenomena, offering a mathematical framework for understanding their behavior. 

In various real-world scenarios, we encounter systems that evolve randomly over time, which can be modeled using martingales. In finance and economics, martingales are an invaluable tool for modeling asset prices \cite{fama1965} and option pricing \cite{Musiela_Rutkowski_2005}. They provide insight into risk assessment, portfolio management, and the efficient market hypothesis, which postulates that asset prices fully reflect all available information \cite{yaes1989}. 

Martingales are also closely related to several important limit theorems in probability theory. These theorems, such as the strong law of large numbers and the central limit theorem, formalize the asymptotic behavior of sample means and sums of random variables. They have profound implications in statistics, allowing us to draw conclusions about large datasets and make predictions based on limited information. Martingale theory allows us to investigate whether these systems remain bounded or converge to certain values in the long run.

In addition to their relevance in mathematics, martingales find applications in various interdisciplinary fields. Their ability to model randomness and analyze dynamic systems makes them useful in physics \cite{roldan2023}, biology, and computer science \cite{mitzenmacher_upfal_2005}, among others.

In the scope of this thesis, we present a formalization of martingales in arbitrary Banach spaces using Isabelle/HOL. We start our discourse by examining existing formalizations in two prominent formal proof repositories, \textsf{mathlib}\footnote{Lean Mathematical Library, which uses the Lean theorem prover based on homotopy type theory.}  and \textsf{AFP}\footnote{Archive of Formal Proofs, a formal proof repository which utilizes  Isabelle/HOL.}. Additionally, we conduct a short review of the literature on conditional expectation and martingales in Banach spaces, laying a solid foundation for our research.

The current formalization of conditional expectation in the Isabelle library is limited to real-valued functions. To overcome this limitation, we extend the construction of conditional expectation to general Banach spaces, employing an approach similar to the one in \cite{Hytoenen_2016}. We compare our construction with this alternative approach in the discussion chapter.

Subsequently, we define stochastic processes and introduce the concepts of adapted, progressively measurable and predictable processes using suitable locale definitions. Most importantly, we provide a generalization for the already present locale \lstinline{filtration} by introducing the locales \lstinline{filtered_measure}, \lstinline{sigma_finite_filtered_measure} and  \lstinline{finite_filtered_measure}. These locales serve to formalize the concept of a filtered measure space. The locale \lstinline{sigma_finite_filtered_measure} also serves to generalize the locale \lstinline{sigma_finite_subalgebra} which is necessary for the development of the theory of martingales. 

Moving forward, we rigorously define martingales, submartingales, and supermartingales, presenting their first consequences and corollaries. Discrete-time martingales are given special attention in the formalization. Overall, we make extensive use of the powerful locale system of Isabelle.

Our formalization fully encompasses the introductory \textsf{mathlib} theory \texttt{probability.mar\-tingale.basic} on martingales, even offering more generalization at certain stages.

The thesis further contributes by generalizing concepts in Bochner integration, extending their application to general Banach spaces. Induction schemes for simple, integrable, and Borel measurable functions on Banach spaces are introduced, accommodating scenarios with or without a real vector ordering. These amendments expand the applicability of Bochner integration techniques.
The thesis concludes with reflections on the formalization approach and suggestions for future research directions.